Cosmic Microwave Background Power Spectra from Few Bit Timestreams

### 2018-08-08 digitisation.tex
Editing.
Done a lot of editing.

\begin{abstract} % copied off ASA for now
Observations of the Cosmic Microwave Background (CMB) are of immense value to modern cosmology. However, future CMB experiments must confront challenges in mission planning, hardware and analysis that arise from the sheer size of the time-ordered-data being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit the data. We investigate using extreme digitisation to address these challenges. Unlike lossless compression, extreme digitisation introduces additional noise into the data. We present an optimal 1, 2 and 3 bit digitisation schemes and lay out how the added noise varies between the schemes and as a function of hits per pixel in the map for temperature and polarisation observations. We show that the noise penalty of 3 bit digitisation is at the percent-level for CMB power spectrum measurements. This is impressive considering that it would reduce the data volume by an order of magnitude. We argue that extreme digitisation is a promising strategy for upcoming experiments.
\end{abstract}

\keywords{ cosmic background radiation --- polarization }
\section{Introduction}
\label{sec:intro}

Observations of the Cosmic Microwave Background (CMB) have played a key role since 1964 (discovery paper). Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will constrain help putting tighter constraints on cosmological models. The most prominent science goal is the discovery of the imprint left by inflationary gravitational waves. Additionally studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects will deliver new insight. CMB experiments however also provide a valuable counterpart to ground-based particle physics experiments, as they probe the relativisitc number of species, the helium fraction and the neutrino mass sum.

The outstanding contribution of CMB science to modern physics has demanded a high standard of data analysis. The CMB community has developed a variety of compression and computational techniques to manage the increasing influx of data, while maximising the science output. These include the compression of time-ordered data into maps, bandpower estimation and the pseudo Cl method.

@@

In this work we present the method of extreme digitisation, which compresses a rich digital input signal and compresses it into a few bits. We apply extreme digitisation to the TOD and detail its effect on temperature and polarisation observations. We find that an optimal 3-bit digitisation scheme adds as little as $\sim 2\%$ to the map noise level. While the digtisation schemes described here are primarily laid out for ground-based Stage-4 experiments, future space-based missions must inevitably study lossy compression techniques.

This work is structured as follows. We detail the arising challenges in handling large TOD in \S\ref{subsec:problem}. We subsequently formulate extreme digitisation formerly in \S\ref{subsec:extremedigitisation} and lay out the framework used to test the performance of this compression technique in \S\ref{subsec:method}. In \S\ref{sec:results} we present the results obtained. We summarize our findings in \S\ref{sec:conclusions}.

@@

The science goals of upcoming CMB experiments naturally lead to a large influx of data. To achieve them longer observations with more detectors are needed. However, the best observations sites for CMB observations are at remote locations: in space and in the Antarctica. Experimetns at these locations depend on satellite transmission The next generation of ground-based experiments, Stage-4, will aim to collect 2 million detector years worth of data. A Stage-4 style experiment will face a data influx of $\sim O(10)Tb/d$. However, the current transmission allocation for the SPT3G is at $150Gb/d$. SPT3G currently deals with its transmission bottleneck by downsampling the TOD, i.e. discarding high frequency information. While this will likely see an increase going into the next stage of experimetns it will not be able to over the entire observation. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of heavy downsampling will narrow the informatino window decisively - prohibiting high l science to be carried out on the transmitted dataset. This also means that any potential faults or errors in the experiment that only become visible in high frequencies will go unnoticed for longer.

Similarly, future space-based missions (PIXIE, COrE+, liteBIRD) aim to increase their detector count by at least an order of magnitude compared to PLANCK. It is questionable whether their telemetry specifications will allow for transmission of the data with PLANCK-style compression. Methods of storing large amounts of data on upcoming satellites will likely be prohibited by financial decisions: the amount of storage space required becomes financially relevant at the scales targeted.

@@

Beyond these transmission hurdles detailed mission planning is becoming exceedingly expensive. As noted by (S4 science book) a full simulation of TOD over the entire parameter space of detection scenarios for numerous set-ups is the desired way to plan Stage-4 configuration. Similarly space-based missions must carry out a similar analysis to optimise their science out-put. However, given the sheer size of TOD this is not possible. We must turn to different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of planning and development stages.

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. Through the expontentially growing size of TOD CMB data analysis is becoming increasingly expensive.

@@

Extreme Digitisation would tackle the challenges mentioned above by recuding the size of the TOD by an order of magnitude. Together with already exploited lossless compression techniques (such as FLAC) this will directly tackle transmission hurdles. While it needs to be investigated to what extent existing algorithms can be carried over, extreme digitisation has the possiblity of solving planning and analysis problems.

@@

Digitisation is a lossy compression technique. However the induced noise depends on the number of bits used, the digitisation thresholds and the output levels chosen. To minimise the noise induced through this process one must know the nature of input signal. A theoretical framework to obtain these levels was laid out by Max in 1978. We review the key aspects of his work relevant to our findings below.

We divide the space $-\infty \to \infty$ into $N$ ranges. One then has endpoints $x_k$ and output levels $y_k$, such that an input between $x_i$ and $x_{i+1}$ produces an output at $y_i$. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$. In order to determine an optimal set of $x_k$ and $y_k$ we would like to minimise the distortion

@@

Seeing as we wish to minimise the distortion we differentiate the above with respect to $x_i$ and $y_i$ and set the derivatives to zero.

\[ \frac{\partial D}{\partial x_i} = \left(x_i-y_{i-1}\right)^2 p(x_i) - \left(x_i - y_i\right)^2 p(x_i) = 0 \]

\[ \frac{\partial D}{\partial x_j} = -2 \int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0 \]

From the first equation above we deduce that

@@

The second one

@@

implies that we should choose $y_i$, such that it halves the area underneath $p(x)$ in the interval $x_i \to x_{i+1}$.

We are interested in the case where our input signal can fairly be approximated by gaussian white noise, i.e. $p(x) = 1/sqrt{2\pi} e^{-x^2/2}$. Even with these assumptions we cannot solve the resulting equations analytically. Instead one can begin by picking $y_1$ and calculating the remaining $x_i$ and $y_i$ using equation \ref{eq:digitequalspacecondition}. One then observes weather this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case the $x_k$ and $y_k$ were chosen appropriately. This is unlikely to succeed on the first try. We rather solve the above system iteratively.

This was carried out by Max. We incorporate his results by formulating the multi-level functions we use for our 1, 2 and 3 bit digitisation process. For 1 bit digitisation we apply the sign function

\[ \hat{x}(t) = \left\{ \begin{array}{lr}
1, & \text{for } x(t) > 0\\
-1, & \text{for } x(t) \leq 0
\end{array} \right. \]

@@

\[ \hat{x}(t) = \left\{ \begin{array}{lr}
1.51 \sigma, & \text{for } x(t) \geq 0.9816 \sigma\\
0.4528 \sigma, & \text{for } 0 \leq x(t) < 0.9816 \sigma\\
-0.4528 \sigma, & \text{for } 0.9816 \sigma \leq x(t) < 0\\
-1.51 \sigma, & \text{for } 0.9816 \sigma < x(t)\\
\end{array} \right. \]

Finally the optimal 3 bit digitisation is described by the eight-level function

\[ \hat{x}(t) = \left\{ \begin{array}{lr}
2.152 \sigma, & \text{for } x(t) \geq 1.748 \sigma\\
1.344 \sigma, & \text{for } 1.05 \sigma \leq x(t) < 1.748 \sigma\\
0.756 \sigma, & \text{for } 0.501 \sigma \leq x(t) < 1.05 \sigma\\
0.245 \sigma, & \text{for } 0 \leq x(t) < 0.501 \sigma\\
-0.245 \sigma, & \text{for } 0.501 \sigma \leq x(t) < 0\\
-0.756 \sigma, & \text{for } 1.05 \sigma \leq x(t) < 0.501 \sigma\\
-1.344 \sigma, & \text{for } 1.748 \sigma \leq x(t) < 1.05 \sigma\\
-2.152 \sigma, & \text{for } 1.748 \sigma < x(t)\\
\end{array} \right. \]

% Theoretical lvls for gaussian white noise

@@

To investigate the performance of the aforementioned digitisation schemes we simulate many scans over CMB template maps at the timestream level. Each scan is performed by a single detector. We obtain maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the additional noise induced through the extreme digitisation process.

The $I, Q, U$ template maps used were created through healpy's synfast. As the input power spectra we use the results by Placnk 2015. The underlying cosmological parameters are summarised in table \ref{tab:inputcosparams}.

We simulate observing a $\sim 600$ \sqdeg patch of the sky. To do so we perform constant elevation scans (CES) over the observation region. We repeat the observation strategy $100$ times with a slight offset in RA and DEC each time, such that all pixels within the path are hit approximately uniformly.

While performing each CES the appropriate pixels that are being targeted are determined. The corresponding values from the template maps are then accessed and added to realisations of the detector noise of appropriate length. We assume the detector noise to be gaussian white noise.

The constructed timestreams for $I$, $Q$ and $U$ are then compressed in 4 different ways: averaging of the timestream, representing each datapoint as a 64bit floating point number and applying the three introduced digitisation schemes before averaging the timestream. The simulation parameters are summarised in table \ref{tab:modelparams}. Through this procedure we slowly build up 12 different maps: three maps that corresponds to a normal scan over the polarised CMB and 9 maps that reconstruct the observed patch from an extremely compressed time-ordered data. Additionally we retain the hitmap.

The simulation described above was implemented in Python 2.7. It has considerable computational requirements if we want to reach up to $\sim 10^8$ hits per pixel. To carry out this simulation we make use of parallelisation and the computing power provided by NERSC.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
\small
\begin{tabular}{l | c c c }
$\Omega$ & XX&XX&$r$\\
\hline

\end{tabular}
\tablecomments{ 
Cosmological parameters used to create the template maps.
} \normalsize
\end{center}
\end{table}

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:modelparams} Assumed Survey Parameters}
\small
\begin{tabular}{c c c c c}
$\mathrm{NSIDE}$ & $f_{readout}$ & $f_{sky}$ & $\sigma^{\mathrm{T}}_{detector}$ & $\sigma^{\mathrm{Pol}}_{detector}$\\
\hline
$4096$ & $200\mathrm{Hz}$ & $\sim 0.014$ & $500\mathrm{K}\sqrt{\mathrm{s}}$ & $\sqrt{2}\times500\mathrm{K}\sqrt{\mathrm{s}}$\\

@@

\subsection{Analysis}
\label{subsec:analysis}

We use PolSpice to compute the $TT, EE, BB$ power spectra of the obtained maps. When doing so we apodise the observed skypatch using a cosine mask to minimise cut sky effects on the spectra. Please see \ref{tab:polspiceparams} for an overview of the parameters used in this process.

@@

The obtained power spectra originating from digitised time streams are normalised against the power spectra that use the full time stream data. We use results that originate from $\sim 10^8hpp$ maps. We separately normalise the TT, EE and BB spectra. The normalisation constant obtained from this calculation are then applied to all lower hit per pixel runs.

We restrict ourselves to windows in multipole space where we trust that the spectra are within appropriate multipole moment range given the patch size and are signal dominated. We therefore place a lower limit on the normalisation window for each channel from

\[ l = \frac{2}{\pi} \frac{(180^\circ )^2}{600 sqdeg} \approx 35 \]

@@

@@

To determine how much the quality of the power spectra suffers due to the digitisation we would like to compare the map noise levels. We would like to put this value into context with the map noise level induced through the detector noise. We therefore formulate

\[ \frac{\Delta \sigma}{\sigma} = \frac{\sigma_{MAP}^{D}-\sigma_{MAP}^{SN}}{\sigma_{MAP}^{SN}} \]

where $\sigma_{MAP}^{SN}$ is the map noise level of the full power spectra (signal plus noise) and $\sigma_{MAP}^{D}$ is the map noise level obtained from the power spectra originating from extremely digitised TOD. We assume that the digitisation process results in adding some constant  to the powerspectrum, i.e.

@@

in the limit where we are dominated by noise. Here $C_l^D$ is the power spectrum originating from a digitised timestream, $C_l^N$ is the detector noise level and $C_l^X$ the additional noise induced. We may now progress with

@@

@@

The results for the 1, 2 and 3 bit digitisation schmes for the TT, EE and BB power spectra are summarised in \ref{fig:extranoise}. We would like to points out three key results. Firstly, 3 Bit digitisation performs the best, followed by 2 Bit and finally 1 Bit digitisation. This is what we expect, given that with each additional bit we may retain more information. Secondly, for a fixed detector noise level, the additional percentage to the map noise level due to digitisation is independent of the number of hits per pixel. This is a valuable result, as it means that there is not a certain point at which surveys start becoming viable for this compression techniques, but they rather perform equally well throughout the range of hits per pixel tested. Lastly, the added noise levels are astonishingly low.

%Any other results that are outstanding

%? Analysis result: To uncertainty in cosmological parameters?

We present the average additional percentage to the map noise level through digitisation in \ref{tab:noise}. These results are obtained by calculating the weighted average across all hits per pixel for a given digitisation scheme and channel. We note that an optimal 3 bit digitisation scheme adds as little as $\sim2\%$ to the map noise level.

Given that advances in CMB observations happen in order magnitude jumps between generations of experiments this is a negligible noise addition, considering that it reduces the TOD by an order of magnitude.

Extreme digitisation is a viable lossy compression technique when dealing with large low signal to noise, slowly varying datasets. Within this limit digitisation relies on the noise to push the small signal towards the correct side of the locally stationary signal. A low signal to noise ratio prevents us from a signal saturating the digitisation levels. For large datasets are needed to provide many samplings of each locally stationary signal. An increase in bit number does not necessarily improve the recovery of the mean, but rather advances narrowing down on its uncertainty, by improving the nulling out of the noise. Improving the number of samples in a dataset improves the finding of the mean, as it allows for more correctly pushed digitised points to be considered.

@@

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:noise} Noise levels}
\small
\begin{tabular}{l | c c c }
 Channel & 1 Bit & 2 Bit & 3 Bit\\
 \hline
 TT & & & \\
 EE & & & \\
 BB & & & \\

\end{tabular}
\tablecomments{ 
Added percentage to the map noise level due to digitisation.
} \normalsize
\end{center}
\end{table}

>>>>

\begin{abstract}
Observations of the Cosmic Microwave Background (CMB) are of immense value to modern cosmology. However, future CMB experiments must confront challenges in mission planning, hardware and analysis that arise from the sheer size of the time-ordered-data being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit the data. We investigate using extreme digitisation to address these challenges. Unlike lossless compression, extreme digitisation introduces additional noise into the data. We present optimal 1, 2 and 3 bit digitisation schemes and determine the degradation in temperature and polarisation power spectra caused by this process. In particular we show that 3 bit digitisation has a percent-level contribution to the map noise level. This is impressive considering that it would reduce the data volume by an order of magnitude. We argue that extreme digitisation is a promising strategy for upcoming experiments.
\end{abstract}

\keywords{ cosmic background radiation --- polarization --- data compression}
\section{Introduction}
\label{sec:intro}

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since 1964 (discovery paper). Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. The most prominent science goal is the discovery of the imprint of inflationary gravitational waves. Additionally studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects will give us access to previously unseen information. Through probing the relativistic number of species, the helium fraction and the neutrino mass sum, CMB experiments are a valuable counter-part to ground-based particle physics experiments.

The outstanding contribution of CMB science to modern physics has demanded a high standard of data analysis. The CMB community has developed a variety of compression and computational techniques to manage the increasing influx of data, while maximising the science output. These include the compression of time-ordered data into maps, bandpower estimation and the pseudo $C_l$ method.

@@

In this work we present the method of extreme digitisation, which compresses a rich digital input signal and compresses it into a few bits. We apply extreme digitisation to the TOD and detail its effect on temperature and polarisation observations. We find that an optimal 3-bit digitisation scheme adds as little as $\sim 2\%$ to the map noise level. While the digtisation schemes described here are primarily laid out for ground-based experiments, future space-based missions must inevitably study lossy compression techniques.

% update this paragraph towards the end
This work is structured as follows. We detail the arising challenges in handling large TOD in \S\ref{subsec:problem}. We subsequently formulate extreme digitisation formerly in \S\ref{subsec:extremedigitisation} and lay out the framework used to test the performance of this compression technique in \S\ref{subsec:method}. In \S\ref{sec:results} we present the results obtained. We summarize our findings in \S\ref{sec:conclusions}.

@@

The science goals of upcoming CMB experiments naturally lead to a large influx of data. To achieve the targeted sensitivity longer observations with more detectors are needed. In fact the number of detectors of ground-based experiments has been following a Moore's law like trend, doubling approximately every 2 years. This directly translates into an exponential growth in data volume.

The best observations sites for CMB measurements are at remote locations: in space and in the Antarctica. Experiments at these locations depend on satellite transmission. The next generation of ground-based experiments will aim to collect 2 million detector years worth of data. An experiment contributing to Stage-4 at the South Pole will face a data influx of $\sim O(10)Tb/d$. However, the current transmission allocation for the SPT3G is at $150Gb/d$, which will likely only see a moderate increase in the coming years. The transmission bottleneck is currently overcome by recovering the full data with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted dataset. This also means that any potential faults or errors in the experiment that only become visible in high frequencies will go unnoticed for longer.

Future space-based missions aim to exceed the detector count of Planck by at least an order of magnitude. It is questionable whether their telemetry specifications will allow for transmission of the data with Planck-style compression. Methods of storing large amounts of data on upcoming satellites will likely be prohibited by financial decisions: the amount of storage space required becomes financially relevant at the scales targeted. Missions will likely be left to design their own compression algorithms which will incorporate a combination of lossless and lossy compression techniques.

@@

Beyond transmission challenges, mission planning is becoming exceedingly expensive. As noted by (S4 science book) a full simulation of TOD over the entire parameter space of detection scenarios for numerous set-ups is the desired way to decide on Stage-4 configurations. Space-based missions must aim to carry out a similar analysis to optimise their science output. We must rely on different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of planning and development stages.

% Analysis

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. Through the exponentially growing size of TOD CMB data analysis is becoming increasingly expensive.

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already exploited lossless compression techniques (such as FLAC) this will directly tackle transmission hurdles. While it needs to be investigated to what extent existing algorithms can be carried over, extreme digitisation has the possibility of solving planning and analysis problems. Other areas of science have already demonstrated that extreme digitisation can be a viable compression technique. For example Jenet and Anderson have shown the advantages of this compression technique to pulsar timing measurements. The computationally challenging searches for continuous gravitational wave searcher may also profit from few-bit digitisation as laid out by Clearwater et al. (in prep.).

@@

Digitisation is a lossy compression technique. However the induced noise depends on the number of bits used, the digitisation thresholds and the output levels chosen. To minimise the noise induced through this process one must know the nature of input signal. A theoretical framework to obtain these levels was laid out by Max in 1978. We review the key aspects of his work relevant for us below.

Digitisation discretises an input signal by sorting it into $N$ appropriate ranges, such that an input between $x_i$ and $x_{i+1}$ produces an output at $y_i$. A digitisation scheme is described the the number of ranges, $N$, the endpoints of these ranges $x_k$ and the output levels $y_k$. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$. In order to quantify the performance of a given digitisation scheme that produces a signal out $s_{out}$ form an input signal $s_{in}$ we define the distortion as

@@

Seeing as we wish to minimise the distortion we differentiate the above with respect to $x_i$ and $y_i$ and set the derivatives to zero. We obtain the two equations

\begin{equation} \label{eq:distderiv1}
\frac{\partial D}{\partial x_i} = \left(x_i-y_{i-1}\right)^2 p(x_i) - \left(x_i - y_i\right)^2 p(x_i) = 0
\end{equation}

\begin{equation} \label{eq:distderiv2}
\frac{\partial D}{\partial x_j} = -2 \int_{x_i}^{x_{i+1}} \left( x-y_i \right) p(x) dx = 0
\end{equation}

Rearranging equation \ref{eq:distderiv1} we deduce

@@

which informs us that an output level $y_i$ must lie halfway between is delimiting thresholds $x_i$ and $x_{i+1}$. We gain an additional condition from equation \ref{eq:distderiv2}

@@

This implies that we should choose $y_i$, such that it halves the area underneath $p(x)$ in the interval from $x_i$ to $x_{i+1}$.

To progress further we have to make an assumption about the distribution of input signals, $p(x)$. For our purposes we may safely assume that ground-based CMB observations operate at low signal to noise. Furthermore we assume that the noise profile is Gaussian white noise\footnote{Please see the conclusion for a discussion of the effect of more realistic noise profile}, i.e. $p(x) = 1/\sqrt{2\pi} e^{-x^2/2}$. Given this assumptions we cannot solve proceed analytically, but can the problem using a numerical iterative procedure. One begins by picking $y_1$ and calculating the remaining $x_i$ and $y_i$ using equation \ref{eq:digitequalspacecondition}. Afterwards one observes weather this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case, the $x_k$ and $y_k$ were chosen appropriately.

This was carried out by Max. We incorporate his results by formulating the multi-level functions we use for our 1, 2 and 3 bit digitisation process. Given an input signal $x(t)$ a digitisation scheme using $N$ bits returns $\hat{x}_N(t)$. For 1 bit digitisation we apply

\[ \hat{x}_1(t) = \left\{ \begin{array}{lr}
1, & \text{for } x(t) > 0\\
-1, & \text{for } x(t) \leq 0
\end{array} \right. \]

@@

\[ \hat{x}_2(t) = \left\{ \begin{array}{rl}
1.51 \sigma, & \text{for } x(t) \geq 0.9816 \sigma\\
0.4528 \sigma, & \text{for } 0 \leq x(t) < 0.9816 \sigma\\
-0.4528 \sigma, & \text{for } 0.9816 \sigma \leq x(t) < 0\\
-1.51 \sigma, & \text{for } 0.9816 \sigma < x(t)\\
\end{array} \right. \]

to the input signal. Finally the optimal 3 bit digitisation is described by the eight-level function

\[ \hat{x}_3(t) = \left\{ \begin{array}{rl}
2.152 \sigma, & \text{for } x(t) \geq 1.748 \sigma\\
1.344 \sigma, & \text{for } 1.05 \sigma \leq x(t) < 1.748 \sigma\\
0.756 \sigma, & \text{for } 0.501 \sigma \leq x(t) < 1.05 \sigma\\
0.245 \sigma, & \text{for } 0 \leq x(t) < 0.501 \sigma\\
-0.245 \sigma, & \text{for } 0.501 \sigma \leq x(t) < 0\\
-0.756 \sigma, & \text{for } 1.05 \sigma \leq x(t) < 0.501 \sigma\\
-1.344 \sigma, & \text{for } 1.748 \sigma \leq x(t) < 1.05 \sigma\\
-2.152 \sigma, & \text{for } 1.748 \sigma < x(t)\\
\end{array} \right. \]

Other digitisation schemes can be thought of that place the digitisation thresholds and output levels in a different way. However, these are not of interest for now, given the ease at which the devised optimal schemes can be implemented and the fact that they will lead to worse performance.

@@

To investigate the performance of the derived digitisation schemes we simulate many scans over CMB template maps at the timestream level. Each scan is performed by a single detector. We obtain maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the additional noise induced through the extreme digitisation process.

To create the template maps we use the healpix framework and the wealth of support available for it. We generate a realisation of I, Q and U maps based on the results of Planck 2015. The key cosmological parameters are summarised in \ref{tab:inputcosparams}.

We simulate observing a $\sim 600$ \sqdeg patch of the sky. To do so we perform a number of equally spaced constant elevation scans (CES) over the observation region. We repeat the observation strategy $100$ times with a slight offset in RA and DEC each time, such that all pixels within the path are hit approximately uniformly. The speed at which we sweep across the survey field is set to a constant to produce a desired number of hits per pixel in the output maps.

While performing each CES the appropriate pixels that are being targeted are determined. The corresponding values from the template maps are then accessed and added to realisations of the detector noise of appropriate length. We assume the detector noise to be Gaussian white noise.

At this point we apply the digitisation schemes to the TOD. We compress the timestream into maps by averaging all hits falling into the same pixel. We produce 13 maps in total: three maps that have not undergone few-bit digitisation at the timestream level for comparison purposes, nine maps with three each corresponding to each digitisation scheme and the hitmap.

The simulation parameters are summarised in table \ref{tab:modelparams}. It has considerable computational requirements if we want to reach up to $\sim 10^8$ hits per pixel. To carry out this simulation we make use of parallelisation and the computing power provided by NERSC.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
\small
\begin{tabular}{l | c}
Parameter & Planck 2015\\
\hline
$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
$H_0$ & $67.26 \pm 0.98$\\
$n_s$ & $0.9652 \pm 0.0062$\\
$\Omega_m$ & $0.316 \pm 0.014$\\
$\sigma_8$ & $0.830 \pm 0.015 $\\
$\tau$ & $0.078 \pm 0.019$\\
$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
\end{tabular}
\tablecomments{ 
Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
} \normalsize
\end{center}
\end{table}

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:modelparams} Assumed Survey Parameters}
\small
\begin{tabular}{l c c c c}
$\mathrm{NSIDE}$ & $f_{\mathrm{readout}} [\mathrm{Hz}]$ & $f_{{\mathrm{sky}}}$ & $\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$\\
\hline
$4096$ & $200$ & $\sim 0.014$ & $500$ & $\sqrt{2}\times500$\\

@@

\subsection{Power Spectrum Estimation}
\label{subsec:psestimation}

We use PolSpice to compute the $TT, EE, BB$ power spectra of the reconstructed maps. When doing so we apodise the observed skypatch using a cosine mask to minimise cut sky effects on the spectra. Please see \ref{tab:polspiceparams} for an overview of the parameters used in this process.

@@

To normalise the the obtained power spectra we focus on the $\sim 10^8 \mathrm{hpp}$ run. We normalise each power spectrum originating from few bit TOD against its corresponding 64bit TOD counterpart. We place a lower limit on the normalisation window in multipole space by considering the size of the observed patch via

\[ l = \frac{2}{\pi} \frac{32400}{600} \approx 35 \]

@@

we close the normalisation window the first instance the above condition is met. The normalisation constants for each digitisation scheme and channel obtained through this way are then applied to all lower hit per pixel simulations.

@@

To quantify how much the quality of the power spectra suffers due to the digitisation compare the map noise levels inferred from the power spectra. These are put into the context of the map noise levels deduced from the 64bit TOD power spectra. We formulate

\begin{equation} \label{eq:extramapnoise}
\frac{\Delta \sigma}{\sigma} = \frac{\sigma_{\mathrm{map}}^{\mathrm{D}}-\sigma_{\mathrm{map}}^{\mathrm{F}}}{\sigma_{\mathrm{map}}^{\mathrm{F}}}
\end{equation}

where $\sigma_{\mathrm{map}}^{\mathrm{F}}$ is the map noise level of the control power spectra and $\sigma_{\mathrm{map}}^{\mathrm{D}}$ is the map noise level obtained from the power spectra originating from extremely digitised TOD. We assume that the digitisation process results in adding some constant independent of multipole moment to the power spectrum. If we are dominated by noise we can write

@@

Here $C_l^D$ is the power spectrum originating from a digitised timestream, $C_l^N$ is the detector noise level and $C_l^X$ the additional noise induced through digitisation. We now progress equation \ref{eq:extramapnoise} to

@@

The value of $l$ at which we can safely assume to be noise dominated and apply the above framework varies between simulated observations of different hits per pixel and for different channels. The range considered involves any datapoints at multipole moments larger than the last point at which the detector noise is at least than 10 times the input power spectrum.

@@

%plot these X/N ?

The results for the 1, 2 and 3 bit digitisation schemes are summarised in figure \ref{fig:extranoise}. We would like to points out three key results. Firstly, 3 Bit digitisation performs the best, followed by 2 Bit and finally 1 Bit digitisation. This is what we expect, given that with each additional bit we may retain more information. Secondly, for a fixed detector noise level, the additional percentage to the map noise level due to digitisation is independent of the number of hits per pixel. The added noise therefore scales in the same fashion as the map noise level with the number of hits per pixel, given a fixed detector noise level. Lastly, the added noise levels are astonishingly low. Keeping in mind that CMB detector sensitivity improves in steps of order of magnitude every few years adding an extra percent level noise term does not deteriorate the results appreciably. This is impressive, given that the use of an optimal 3 bit digitisation scheme will save approximately an order of magnitude in TOD volume.

Extreme digitisation is a viable lossy compression technique when dealing with large volume, low signal to noise, slowly varying datasets. Making these assumptions the signal small and constant for a considerable number of datapoints. For such an interval the many samplings of the noise result in producing appropriate numbers in each output bin. The addition of more bits helps by classifying the size of the noise more and more appropriately and hence approaching a good estimate for the value of the signal faster. It is necessary to remain in the low signal to noise regime, as otherwise a dominant signal would saturate the digitisation scheme and for a given scheme only values in the range $\pm |N_{\mathrm{hits}}y_N|$ can be represented, given that we have $N_{\mathrm{hits}}$ datapoints. We must demand the data to be slowly varying such that many datapoints can be used to reconstruct a single significant step of the input signal. Finally a large data volume is clearly favourable for the reasons illustrated above: it allows for repeated sampling and therefore more accurate reconstruction of the weak signal.

@@
